Synthesizing Quotient Lenses

### 2017-10-27 papers/QuotientLensSynth/main.tex
Made changes to QRE section

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; (#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\mathit{get}}}
\newcommand{\lput}{\ensuremath{\mathit{put}}}
\newcommand{\create}{\ensuremath{\mathit{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space = project wsp+ $\to$ " "
let wsp_to_space_epsilon = project wsp* $\to$ ""
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
let wsp_to_comma = wsp_to_space_epsilon . "," . wsp_to_space
let comma_name = name | (name . wsp_to_comma . (name . wsp_to_space)* . name)
let space_to_comma = synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")}
let valid_name = squash comma_name -> space_name using (get space_to_comma)
let wsp_to_and = project (wsp+ . "and" . wsp+) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp+)
$\to$ ",\n") let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

\subsection{Overview}

We have already encountered the \cd{project R} $\to$ \cd{s} QRE
which sends all elements of the language described by the regular expression
$R$ to the canonical representative $s$, and the

\noindent \cd{perm (q_1, ..., q_n) with q} QRE which sends all permutations of
the QREs $q_1, \ldots, q_n$ separated by the QRE $q$ to the canonical
QRE $q_1 \cdot q \cdot \ldots \cdot q \cdot q_n$.

Now consider a variant of the \bibtex{} to EndNote example transformation,
where, the only quotienting occurs in the author field, where the author's name
can have the last name followed by a comma and then the rest of the name, or
remain as before where the names are separated by a space.

For example, these two records should be considered equivalent:
\begin{verbatim}
@Book {conway,
  Author = "Conway, J. H.",
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

The equivalence described is one in which there is a bijection between 
the regular expressions \cd{comma_name} and \cd{space_name} given by,
\begin{lstlisting}
let comma_name : regexp = name | (name . ", " . name . (" " . name)*)

let space_name : regexp = name | (name . " " . (name . wsp_to_space)* . name)

\end{lstlisting}

\noindent with the lens \cd{comma_to_space} mapping between them given by 
\begin{lstlisting}
let comma_to_space : (lens in space_name <=> comma_name) =
lens_swap 
(name . del " " . (ins " " . name . del " ")* ) (name . ins ", ") || name 
\end{lstlisting}

The $\squash{R}{R'}{f}$ QRE is designed to make it easy to express such
equivalences easily. Specifically, given regular expressions $R$ and $R'$, and
a function $f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$, the
$\squash{R}{R'}{f}$ QRE expresses the equivalence relation on $\mathcal{L}(R \;
R')$ which sends each element $r \in \mathcal{L}(R)$ to the canonical element
$f(r) \in \mathcal{L}(R')$, and sends each element $r' \in \mathcal{L}(R')$ to
itself:

\begin{lstlisting}
let squash_author = squash space_name $\to$ comma_name using comma_to_space

let author : canonizer = "author = \"" . squash_author . "\""
\end{lstlisting}

In this example, the \cd{squash_author} QRE expresses the fact that the
\cd{space_name} and \cd{comma_name} representations of the authors' name
are equivalent, with the \cd{comma_name} representation chosen as the
canonical representation, and the lens \cd{comma_to_space} giving
the canonizing function which converts data in the \cd{space_name} format
to its \cd{comma_name} representation.

The $\squash{R}{R'}{\ell}$ has the added advantage that if the transformation
from $R$ to $R'$ can be expressed by a bijective lens that is consistent with a set
of examples, then the programmer can synthesize the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:

\begin{lstlisting}
let comma_to_space : (lens in space_name <=> comma_name) = 
synth space_name <=> comma_name using {("J. H. Conway", "Conway, J. H.")}
\end{lstlisting}

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} record is named
\cd{bibtex}, then the QRE for matching two records is 
\cd{bibtex . bibtex}, the QRE for matching one or two records is
\cd{ bibtex | (bibtex . bibtex)} and the QRE for matching any number of QRE
records \cd{bibtex*}

The $\normalize{R}{R'}{f}$ expresses the equivalence relation which sends each
element in $\mathcal{L}(R)$ to its canonical representation $f(r) \in
\mathcal{L}(R')$. The equivalence relation defined by
$\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QREs
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations which are
difficult to express, or which cannot be expressed, using the other
combinators. However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

@@

In summary each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.
\subsection{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep R \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\project}[2]{\ensuremath{\mathit{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; #1 \rightarrow #2
\mathit{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\mathit{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\mathit{put}}}
\newcommand{\create}{\ensuremath{\mathit{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space = project wsp+ $\to$ " "
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
let comma_name = name | (name . "," . wsp_to_space . (name . wsp_to_space)* . name)
let space_to_comma = synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")} 
let valid_name = squash comma_name -> space_name using (get space_to_comma)
let wsp_to_and = project (wsp+ . "and" . wsp+) $\to$ " and " 
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}" 
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep \project{R}{s} \sep \squash{q}{q'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q \semicolon q' \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

\subsection{Semantics of Quotient Regular Expression}
In the \bibtex{} to EndNote example, we encountered the $\project{R}{s}$,

\noindent $\squash{q}{q'}{f}$ and $\perm{q_1, \ldots, q_n}{q}$ QREs, as well as
the regular operators applied to these, as in

\begin{lstlisting}
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)
\end{lstlisting}

The $q \semicolon q'$ QRE is a compostion combinator used to successively apply
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:

\begin{lstlisting}
(* Define a QRE which matches sends all comma names to a space name equivalent *)
let wsp = [ \n\t\r] 
let whole_space_name  = name | (name . wsp+ . (name . wsp+)* . name) 
let whole_comma_name = name | (name . "," . wsp+ . (name . wsp+)* . name) 
let whole_space_to_comma = 
synth whole_comma_name <=> whole_space_name using {("Conway, J. H.", "J. H. Conway")} 
let whole_valid_name = 
squash whole_comma_name -> whole_space_name using (get whole_space_to_comma)

(* Define QRE that sends space names to their canonical representation *)
let wsp_to_space = project wsp+ $\to$ " "
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)

(* Compose the two QREs *)
let valid_name = whole_valid_name ; space_name
\end{lstlisting}

The final combinator is the  $\normalize{R}{R'}{f}$ combinator, which expresses
the equivalence relation which sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QREs
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations which are
difficult to express, or which cannot be expressed, using the other
combinators.

However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

@@

In summary each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.

Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $q, q'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(q) \cdot W(q')$, $W(q) \sep W(q')$,
$W(q)^*$, $K(q) \cdot K(q')$, $K(q) \sep K(q')$, $K(q)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.
